Hydraulically-vulnerable trees survive on deep-water access during droughts in a tropical forest

### 2021-04-09 10.0_Supporting_Information.Rmd
Explained uncertainty in isotopic data

\newpage
```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic ratio ($\\mathrm{\\delta}^2H$) of soil water by depth measured at Barro Colorado Island in March 1997. Data from @Meinzer:1999."}
knitr::include_graphics(file.path("figures/UDI_confidence/Meinzer_etal_1999_Depth_vs_soil_deltaD.jpeg"))
```

@@

For each ERD model structure (Eqs. \@ref(eq:M1)-\@ref(eq:M6)), species-specific best-fit ERD was identified as follows: For example, for the chosen model (Eq. \@ref(eq:M3)), for a species $s$ and hydrological realization *h*, we obtained Akaike's Information Criterion (AIC) and Goodness-of-fit ($R$^2^) for each modeled soil layer depth $z$. We used AIC as the first selection criterion to ensure that the fitted growth values from Eq. \@ref(eq:M3) are within the confidence interval of the error distribution of observed growth (as the same model Eq. \@ref(eq:M3) was being compared for each *h*, number of model parameters were the same). For each *h*, we retained those depths with the maximum likelihood of growth (similar AIC based support, see below) filtering out the rest. However, AIC does not ensure that fitted values will have a positive trend with growth. We therefore used $R$^2^ between observed and fitted growth values as a goodness-of-fit measure and retained depths associated with a relatively high threshold $R$^2^ value of 0.5; after filtering out depths associated with negative correlations $r$ if any. Finally, among these we chose the depth with the highest $R$^2^ per *h*. Results were not sensitive to inclusion of this final step of retaining only a single depth as the best-fit (not shown). For a species $s$, we defined effective rooting depth (ERD) as the median of the best-fit depth over all *h*, which can be described as:

@@

>>>>

\newpage
```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic ratio ($\\mathrm{\\delta}^2H$) of soil water by depth measured at Barro Colorado Island in March 1997. Data from @Meinzer:1999. $\\mathrm{\\delta}^2H$ to depth relationship was particularly uncertain for $\\mathrm{\\delta}^2H$ values > -40 ‰. Two species, *Guapira standleyana* (-28.9 $\\pm$ 3.7; mean $\\pm$ SE) and *Spondias radlkoferi* (-33.5 $\\pm$ 2.6), in @Meinzer:1999 dataset had $\\mathrm{\\delta}$^2^H~xylem~ values > -40 ‰. For example, *Guapira standleyana*, with $\\mathrm{\\delta}$^2^H value of -28.9 $\\pm$ 3.7 could have a source-water depth anywhere between 10 to 50 cm. *Spondias radlkoferi* is also typically leafless in March-April. These two species were not included in the ERD model-data comparison."}
knitr::include_graphics(file.path("figures/UDI_confidence/Meinzer_etal_1999_Depth_vs_soil_deltaD.jpeg"))
```

@@

For each ERD model structure (Eqs. \@ref(eq:M1)-\@ref(eq:M6)), species-specific best-fit ERD was identified as follows: For example, for the chosen model (Eq. \@ref(eq:M3)), for a species $s$ and hydrological realization *h*, we obtained Akaike's Information Criterion (AIC) and Goodness-of-fit ($R$^2^) for each modeled soil layer depth $z$. We used AIC as the first selection criterion to ensure that the fitted growth values from Eq. \@ref(eq:M3) are within the confidence interval of the error distribution of observed growth (as the same model Eq. \@ref(eq:M3) was being compared for each *h*, number of model parameters were the same). For each *h*, we retained those depths with the maximum likelihood of growth (similar AIC based support, see below) filtering out the rest. However, AIC does not ensure that fitted values will have a positive trend with growth. We therefore used $R$^2^ between observed and fitted growth values as a goodness-of-fit measure and retained depths associated with a relatively high threshold $R$^2^ value of 0.5; after filtering out depths associated with negative correlations $r$, if any. Finally, among these we chose the depth with the highest $R$^2^ per *h*. Results were not sensitive to inclusion of this final step of retaining only a single depth as the best-fit (not shown). For a species $s$, we defined effective rooting depth (ERD) as the median of the best-fit depth over all *h*, which can be described as:

@@

Deciduous leaf-habits are generally expected to be fully deciduous at individual tree-level. At BCI, however, there is a great within-species variation in the timing of leaf-drop and leaf-out, and neither do all individuals of a deciduous species reach full deciduousness every year (Condit et al., 2000). Our reconstruction of LAI seasonality is based upon species-level leaf-fall data, which captures the within-species variation in the timing of leaf-drop, so in our reconstruction seasonal decline in $\textrm{LAI}^*$ appears to be more spread out and does not reach zero, in contrast to what may be  expected for an idealised individual of each deciduous leaf habit. Greater uncertainty in our estimates of leaf-lifespan for deciduous species may have compounded this issue further. We chose not to normalise the deciduous species’ $\textrm{LAI}^*$ curves to reach zero as it would further distort the shape of the curve, shifting a larger period under a lower $\textrm{LAI}^*$.
